% !TEX root = master.tex
% !TEX program = lualatex
\lecture{4}{12-03-2025}{Laurent series and Taylor series}{}

\subsection{Remark: trick to compute contour integrals}

A useful trick to compute contour integrals is the following.

\begin{itemize}
\item Find another simple closed curve that surrounds the same point as the complicated curve.
\item If the function is holomorphic on the region between the two curves, then the integrals along the two curves are equal.
\end{itemize}

In other words, if $\Gamma_0$ and $\Gamma_1$ are two curves with the same endpoints and orientation, and if $f$ is holomorphic on the region between them, then

\[
\int_{\Gamma_0} f(z)\,dz
=
\int_{\Gamma_1} f(z)\,dz .
\]

The idea is to apply Cauchy's theorem to the closed contour formed by $\Gamma_0$ and $-\Gamma_1$.

\subsection{Extended Cauchy Integral Formula}

Under the same assumptions, for every integer $n \ge 0$ we have

\begin{corollary}
\[
f^{(n)}(z_0)
=
\frac{n!}{2\pi i}
\int_\Gamma
\frac{f(z)}{(z-z_0)^{n+1}}\,dz .
\]

 
\end{corollary}
Equivalently,

\[
\int_\Gamma
\frac{f(z)}{(z-z_0)^{n+1}}\,dz
=
\frac{2\pi i}{n!}f^{(n)}(z_0).
\]

\textbf{Remark.}  
This formula shows that holomorphic functions are infinitely differentiable.

\subsection{Examples}

\subsubsection*{Example 1}

Let

\[
f(z)=\frac{\cos z}{z}
\]

and let $\Gamma$ be a positively oriented simple closed curve.

We distinguish three cases:

\begin{itemize}

\item If $0$ lies on $\Gamma$, the integral is not defined.

\item If $0$ is outside $\Gamma$, then

\[
\int_\Gamma \frac{\cos z}{z}\,dz = 0
\]

by Cauchy's theorem.

\item If $0$ is inside $\Gamma$, then using Cauchy's integral formula

\[
\int_\Gamma \frac{\cos z}{z}\,dz
=
2\pi i \cos(0)
=
2\pi i .
\]

\end{itemize}

\subsubsection*{Example 2}

Let $\Gamma = \{z \in \mathbb{C} : |z-2| = 1\}$ oriented counterclockwise.  
Compute

\[
\int_\Gamma \frac{e^z}{(z-2)^3}\,dz .
\]

Using the extended Cauchy formula with $f(z)=e^z$, $z_0=2$ and $n=2$,

\[
\int_\Gamma \frac{e^z}{(z-2)^3}\,dz
=
\frac{2\pi i}{2!} f^{(2)}(2).
\]

Since

\[
f^{(2)}(z_0)=e^2,
\]

we obtain

\[
\int_\Gamma \frac{e^z}{(z-2)^3}\,dz
=
\frac{2\pi i}{2}e^2
=
\pi i e^2.
\]
\newpage 

\chapter{Laurent series and Taylor series}\label{chap:laurent_series_and_taylor_series} % (fold)

\section{Taylor Polynomial}

Let $f : \mathbb{R} \to \mathbb{R}$ be sufficiently regular (N-times differentiable).  
For any $x_0 \in \mathbb{R}$, the Taylor polynomial of degree $N$ is

\[
T^N_{x_0}f(x)
=
f(x_0)
+
f'(x_0)(x-x_0)
+
\cdots
+
\frac{f^{(N)}(x_0)}{N!}(x-x_0)^N.
\]

\subsection{Taylor's Theorem}

If $f \in C^{N+1}$, then for all $x_0$, $x \in \mathbb{R}$ there exists $\vartheta \in (0,1)$ such that

\[
f(x)
=
T^N_{x_0}f(x)
+
\frac{f^{(N+1)}(\vartheta x_0+(1-\vartheta)x)}{(N+1)!}(x-x_0)^{N+1}.
\]

\textbf{Remark.}  
Taylor's theorem shows that a function can be approximated locally by a polynomial.

\subsection{Taylor Series}

If $f$ has derivatives of all orders, we can define the Taylor series

\[
\sum_{n=0}^{\infty}
\frac{f^{(n)}(x_0)}{n!}(x-x_0)^n.
\]

A natural question is whether this series converges and whether it converges to the function $f$.

\subsection{Example}

For the exponential function,

\[
e^x
=
\sum_{n=0}^{\infty}
\frac{x^n}{n!}.
\]

This series converges for all $x \in \mathbb{R}$. \\
\textbf{Remark:} Functions $f$ s.t the Taylor series converges to $f$ in a neighborood of $x_o$ are called \textbf{analytic}. (Not all smooth (continually derivable) functions are analytic)

\subsection{Taylor coefficients}
\begin{definition}[Taylor Coefficients]
  The Taylor coefficients of a function $f(z)$ around a point $z_0$ are defined as: \[
    c_n = \frac{f^{(n)}(z_0)}{n!}
  \]
  where $f^{(n)}(z_0)$ is the $n$-th derivative of $f$ evaluated at $z_0$.

  This comes from assuming $f$ has a power series representation: \[
    f(z) = \sum_{n=0}^{\infty}c_n(z-z_0)^n  
  \]
  
\end{definition}
\section{Taylor series / regular Laurent series for complex functions}

\begin{theorem}[]
  Let $D \subset \mathbb{C}$ be open and let $z_0 \in D$.  Let $R>0$ such that, the disk $B_R(z_0) = \{z \in \mathbb{C} : \left|z-z_0\right| < R\}$ is contained in D.

If $f : D \to \mathbb{C}$ is holomorphic.

Then for all $z \in B_R(z_0)$

\[
f(z)=\sum_{n=0}^{\infty} \frac{f^{\left(n\right)}(z_0)}{n!} (z-z_0)^n
\]
\end{theorem}

The largest possible $R$ is called \textbf{radius of convergence} (It's the distance between $z_0$ and the nearest singularity (not $z_0$).

\textbf{Remark:}

$(1)$ All holo functions are analytic. 

$(2)$ In practice: $R$ is the distance of $z_0$ from the $nearest$ point where $f$ is not holo or not defined. 

$(3)$ We can connect the Taylor series to Cauchy's integral formula \[
  f(z) = \sum_{n = 0}^{\infty} \frac{f^{\left(n\right)}(z_0)}{n!}(z-z_0)^n = \sum_{n=0}^{\infty}c_n(z-z_0)^n 
\]
where \[
  c_n = \frac{f^{\left(n\right)}(z_0)}{n!} = \frac{1}{2 \pi i}\int_{\gamma} \frac{f(\zeta)}{\zeta - z_0} d\zeta
\]
for every simple, close curve $\gamma$ aroung $z_0$ with int $\gamma \subset D$ and counterclockwise oriented.

\textbf{Caution:} The Taylor series representation around a given point is unique! In particular in order to compute the Taylor series at a given point, we can use other "tricks".

\begin{theorem}[Laurent series]
 Let $D \subset \mathbb{C}$ be open and let $z_0 \in D$.  
Assume $f : D \setminus \{z_0\} \to \mathbb{C}$ is holomorphic.

Let $R>0$ such that $B_R(z_0) \subset D$.

Then for all $z \in B_R(z_0)\setminus\{z_0\}$

\[
f(z)=\sum_{n=-\infty}^{\infty} c_n (z-z_0)^n
\]

where for every simple closed counterclockwise oriented curve $\gamma$
such that $\mathrm{int}(\gamma)\subset D$ and $z_0$ is enclosed by $\gamma$,

\[
c_n=\frac{1}{2\pi i}\int_\gamma
\frac{f(\zeta)}{(\zeta-z_0)^{n+1}}\,d\zeta .
\]

\end{theorem}
\paragraph{Remark}

\begin{itemize}

\item The Laurent series is like a Taylor series but it may include
terms with polynomial singularities (negative powers).

\item As for Taylor series, the Laurent series representation is unique.
Hence if we determine the coefficients $c_n$ by any method
(not necessarily by path integrals), we have determined the Laurent series.

\item We do not assume that $f$ is holomorphic at $z_0$.
It might be holomorphic there or not; no hypothesis is made.

\end{itemize}

\begin{definition}[]
 Any Laurent series can be split into a \textbf{singular part} and a
\textbf{regular part}:

\[
\sum_{n=-\infty}^{\infty} c_n (z-z_0)^n
=
\underbrace{\sum_{n=1}^{\infty} c_{-n}(z-z_0)^{-n}}_{\text{singular part}}
+
\underbrace{\sum_{n=0}^{\infty} c_n (z-z_0)^n}_{\text{regular part}} .
\]

 
\end{definition}
In other words, a Taylor series is a Laurent series whose singular
part vanishes.

\paragraph{Example}

If $f : D \to \mathbb{C}$ is holomorphic and $z_0 \in D$,
then its Taylor expansion

\[
f(z)=\sum_{n=0}^{\infty} c_n (z-z_0)^n
\]

is a Laurent series whose singular part is zero since for all $n\leq -1$\[
  \frac{f(\zeta)}{(\zeta-z_0)^{n+1}} \text{is holo. near} z_0
\]
so by Cauchy's integral formula $\int_{\gamma} \frac{f(\zeta)}{(\zeta-z_0)^{n+1}} d\zeta = 0$
\paragraph{Example}

For the following function $f: \mathbb{C}\setminus{0} \to \mathbb{C}$ let find the Laurent series around $z_0 = 0$


1) $f(z)=\frac{1}{z}$, this is already a Laurent serieswith
\[
c_{-1}=1, \qquad c_n=0 \ \text{otherwise}.
\]

2) $f(z) = \frac{5z + 1}{z^2} = \frac{1}{z^2} + \frac{5}{z} = \sum_{n=-\infty}^{\infty}c_nz^n$ where \[
  c_n = 
  \begin{cases}
    1 \quad \text{if }n = -2 \\
    5 \quad \text{if }n = -1 \\
    0 \quad otherwise
  \end{cases}
\]
